---
title: "Introduction to package 'precautionary'"
author: "David C. Norris"
date: "7/8/2020"
output: rmarkdown::html_vignette
vignette: <
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to package 'precautionary'}
  \usepackage[utf8]{inputenc}
bibliography: precautionary-package.bib
---

```{r setup, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE) # suppress un-needed warning
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 5)

library(precautionary)
library(knitr)
```

## Introduction to package `precautionary`

In package [escalation](https://CRAN.R-project.org/package=escalation), you can simulate a 3 + 3 design as follows:

```{r sim-a-3plus3}
library(escalation)
# Design a 3 + 3 trial, with 5 pre-specified doses to escalate through
design <- get_three_plus_three(num_doses = 5)
# Posit a scenario where these 5 doses cause dose-limiting toxicities
# (DLTs) in 12%, 27%, etc. of the population:
scenario <- c(0.12, 0.27, 0.44, 0.53, 0.57)
design %>% simulate_trials(   # Feed the design to the simulator ...
    num_sims = 100            # run 100 simulated trials
  , true_prob_tox = scenario  # under the chosen scenario,
  ) -> sims                   # and store the results.

kable(summary(sims)) # summarize results
```

From an exercise such as this, we learn a few things of interest. For example, we see that, with the lowest dose being too toxic for 12% of the population, in this scenario there is a nonzero probability that even this dose will be rejected by our trial. Conversely, we also find it is quite possible the trial could recommend a dose that is toxic to nearly half the population. Note that, by summing the $n$ column, we can estimate expected enrollment at `r sum(summary(sims)$n)`, possibly informative about the expected cost or duration of our trial. Any complete simulation study will of course consider multiple scenarios, perhaps weighted according to their likelihood.

But simulations such as these cannot answer crucial questions about **trial safety**, mainly because the simulation machinery recognizes no notion of *graded toxicity*. The binary (yes/no) toxicities in the simulation machinery of `escalation` cannot distinguish Grade 5 (fatal) toxicities from Grade 3. 

## Introducing graded toxicities via package `precautionary`

The `precautionary` package solves this problem like this:

```{r introducing-u_i}
library(precautionary)
# A realistic approach to dose-toxicity relations does not begin
# by pulling a sequence of toxicity probabilities out of thin air.
# Rather, it will DERIVE these probabilities according to how a
# latent toxicity threshold is distributed in the population.
mtdi_dist <- mtdi_lognormal(CV = 0.5        # coefficient of variation
                           ,median = 6      # median DLT threshold
                           ,units = "mg/kg" # real doses have units!
                           )

# Likewise, the pre-specified doses for our dose-escalation design
# must be given in ACTUAL dose units. With package 'precautionary',
# this is accomplished by setting a 'dose_levels' option:
options(dose_levels = c(0.5, 1, 2, 4, 6)) # specify actual dosing

# Working within this minimally modified set-up, the 'precautionary'
# package simulates graded toxicities by applying a user-specified
# transformation of the dose-space, termed an 'ordinalizer':
mtdi_dist@ordinalizer <- function(dose, r0) {
   c(Gr1=dose*r0^2, Gr2=dose*r0, Gr3=dose, Gr4=dose/r0, Gr5=dose/r0^2)
}

design %>% simulate_trials(
  num_sims = 100
, true_prob_tox = mtdi_dist  # No mere list of probs now, but a distribution!
, r0 = 1.5                   # Pass this r0 parameter on to the ordinalizer
) -> sims_safety

kable(summary(sims_safety))
```

```{r safety}
kable(attr(summary(sims_safety),'ordtox'))
```

### Fit a hyper_mtdi_distribution to a CRM skeleton

One way to understand the `hyper_mtdi_distribution` class is as a putative generator for the 'skeleton' of a CRM model.

#### Obtain dose-wise DLT probabilities from an MTDi distribution:

Use a lognormally-distributed MTD$_i$ inspired by [@norris_retrospective_2020]:

```{r derive-DLT-probs, eval=FALSE}
mu <- 5              # Values estimated in arXiv:2004.12755 [stat, q-bio];
sigma <- 1/sqrt(1.3) # see subsection 'Model estimates' and Table 2, pp. 2-3.
dose_levels <- c(2, 6, 20, 60, 180, 400)
MTDi <- Lognormal$new(meanlog = mu, sdlog = sigma)
DLT_probs <- MTDi$cdf(dose_levels)
```

#### Run a simple simulation

```{r simple-sim, eval=FALSE}
suppressWarnings( # "Unknown or uninitialised column: `u_i`" warnings (from Tibble?)
sims <- get_three_plus_three(num_doses = 6) %>%
  simulate_trials(num_sims = 200, true_prob_tox = DLT_probs)
)
```

#### Obtain ordinal toxicity outcomes via `u_i`


#### Try accelerated titration à la NCT02848911

TODO.
