---
title: "Introduction to package 'precautionary'"
author: "David C. Norris"
date: "7/8/2020"
output: rmarkdown::html_vignette
vignette: <
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to package 'precautionary'}
  \usepackage[utf8]{inputenc}
bibliography: precautionary-package.bib
---

```{r setup, include=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE) # suppress un-needed warning
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 4, fig.width = 6)

library(precautionary)
library(knitr)
```

## Background

In package [escalation](https://CRAN.R-project.org/package=escalation), you can simulate a 3 + 3 design as follows:

```{r sim-a-3plus3}
library(escalation)
# Design a 3 + 3 trial, with 5 pre-specified doses to escalate through
design <- get_three_plus_three(num_doses = 5)
# Posit a scenario where these 5 doses cause dose-limiting toxicities
# (DLTs) in 12%, 27%, etc. of the population:
scenario <- c(0.12, 0.27, 0.44, 0.53, 0.57)
design %>% simulate_trials(   # Feed the design to the simulator ...
    num_sims = 100            # run 100 simulated trials
  , true_prob_tox = scenario  # under the chosen scenario,
  ) -> sims                   # and store the results.

kable(summary(sims)) # summarize results
```

From an exercise such as this, we learn a few things of interest. For example, we see that, with the lowest dose being too toxic for 12% of the population, in this scenario there is a nonzero probability that even this dose will be rejected by our trial. Conversely, we also find it is quite possible the trial could recommend a dose that is toxic to nearly half the population. Note that, by summing the $n$ column, we can estimate expected enrollment at `r sum(summary(sims)$n)`, possibly informative about the expected cost or duration of our trial. Any complete simulation study will of course consider multiple scenarios, perhaps weighted according to their likelihood.

But simulations such as these cannot answer crucial questions about **trial safety**, mainly because the simulation machinery recognizes no notion of *graded toxicity*. The binary (yes/no) toxicities in the simulation machinery of `escalation` regard Grade 5 (fatal) toxicities no differently from Grade 3. 

## Introducing realistic pharmacologic thinking

The `precautionary` package solves this problem by pursuing a *realistic* approach to pharmacologic thinking. Rather than plucking a sequence of toxicity probabilities out of thin air, this approach **derives** such probabilities according to how a latent toxicity threshold is distributed in the population.

```{r}
library(precautionary)
mtdi_dist <- mtdi_lognormal(CV = 2        # coefficient of variation
                           ,median = 5      # median DLT threshold
                           ,units = "mg/kg" # real doses have units!
                           )
```

Likewise, the pre-specified doses for our dose-escalation design must be specified as actual doses. With package `precautionary`, this is accomplished by setting a `dose_levels` option:

```{r}
options(dose_levels = c(0.5, 1, 2, 4, 6)) # specify actual dosing
```

A plot of the MTD$_i$ distribution makes clear the connection with toxicity probabilities:

```{r}
plot(mtdi_dist)
```

```{r}
mtdi_dist %>%
  tox_probs_at(getOption('dose_levels'), r0 = 1.5) %>%
  t %>% kable(digits = 3)
```

## Introducing graded toxicities via package `precautionary`

The latent toxicity threshold MTD$_i$ introduced here has far-reaching implications for the design of dose-finding trials. Indeed, it forms the conceptual basis for dose-*titration* designs that abandon cohortwise dose-escalation altogether [@norris_dose_2017,@norris_precautionary_2017].

But for *present* purposes---that is, taking it for granted that we must (for whatever reason) employ a dose-escalation design---the MTD$_i$ concept *by itself* amounts to little more than a formality. But in conjunction with a user-specified dose scaling that links different toxicity grades at the individual level, the MTD$_i$ becomes an effective tool for exploring ordinal toxicities even within the confines of a dose-escalation design. In package `precautionary`, such scaling functions are called 'ordinalizers', and are attached to the MTD$_i$ distributions themselves:

```{r}
mtdi_dist@ordinalizer <- function(dose, r0) {
  toxdose_scaling <- r0 ^ c(Gr1=-2, Gr2=-1, Gr3=0, Gr4=1, Gr5=2)
  return(dose / toxdose_scaling)
}
```

To get a sense for what this ordinalizer means, suppose an individual has an MTD$_i$ located at the 30th percentile:

```{r}
MTD30 <- mtdi_dist@dist$quantile(0.30)
mtdi_dist@ordinalizer(MTD30, r0=2)
```

WOW! This shows how untenable this explanation becomes, with the ordinalizer defined in this way. I need to figure out a way to compute the inverse of the specification!

```{r}
design %>% simulate_trials(
  num_sims = 100
, true_prob_tox = mtdi_dist  # No mere list of probs now, but a distribution!
, r0 = 1.5                   # Pass this r0 parameter on to the ordinalizer
) -> sims_safety

kable(summary(sims_safety))
```

```{r safety}
kable(attr(summary(sims_safety),'ordtox'))
```

### Details on the 'ordinalizer'

The ordinalizer given above adopts the basic assumption that the ordinal toxicities are equally spaced in logarithmic dose-space. For example, if $r_0 = 2$, then ...

### Fit a hyper_mtdi_distribution to a CRM skeleton

One way to understand the `hyper_mtdi_distribution` class is as a putative generator for the 'skeleton' of a CRM model.

#### Obtain dose-wise DLT probabilities from an MTDi distribution:

Use a lognormally-distributed MTD$_i$ inspired by [@norris_retrospective_2020]:

```{r derive-DLT-probs, eval=FALSE}
mu <- 5              # Values estimated in arXiv:2004.12755 [stat, q-bio];
sigma <- 1/sqrt(1.3) # see subsection 'Model estimates' and Table 2, pp. 2-3.
dose_levels <- c(2, 6, 20, 60, 180, 400)
MTDi <- Lognormal$new(meanlog = mu, sdlog = sigma)
DLT_probs <- MTDi$cdf(dose_levels)
```

#### Run a simple simulation

```{r simple-sim, eval=FALSE}
suppressWarnings( # "Unknown or uninitialised column: `u_i`" warnings (from Tibble?)
sims <- get_three_plus_three(num_doses = 6) %>%
  simulate_trials(num_sims = 200, true_prob_tox = DLT_probs)
)
```

#### Obtain ordinal toxicity outcomes via `u_i`


#### Try accelerated titration à la NCT02848911

## References
